---
title: "Social support network structure determines resilience to personal disasters"
author: "Richard Morris"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      eval=F,
                      fig.path = "figures/")
library(tidyverse)
```

## Introduction

This is an extension of our paper titled ["The Differential Impact of Major Life Events on Cognitive and Affective Wellbeing"](https://www.sciencedirect.com/science/article/pii/S2352827319302204). That paper described the average response to different life-events described in [HILDA](https://melbourneinstitute.unimelb.edu.au/hilda), with the worst events being _divorce, bankruptcy or loss of a spouse_. There we saw that the average response to the worst life events was a resilient u-shaped response in happiness levels, i.e., on average, happiness recovered to the prior baseline.

This report takes the three worst life-events identified in our previous work (*bankruptcy, divorce, loss*), and identifies three different response-types to those events, i.e., subgroups within the average response, determined by parametric analysis: 1) **resilient**, representing the average response (u-shaped); 2) **vulnerable**, representing a decreasing change in happiness around or after the event; or 3) **relieved**, representing a positive change in happiness around or after the event.  

##### Figure 1. Response types to bad events
```{r patterns, fig.height = 2.75, eval=T}
par(mfrow=c(1,3))
curve(1*x, from=-1, to=1, ylim = c(-1, 1),
      xlab="", ylab="", main = "relief")
curve(-1*x, from=-1, to=1, ylim = c(-1, 1),
      xlab="", ylab="", main = "vulnerable")
curve(x^2 - 0.5, from=-1, to=1, ylim = c(-1, 1), 
      xlab="", ylab="", main = "resilient")
```

Figure 1 above shows the three different response types to life-events.  

We examine the **social support networks** among the three different subgroups of people which display each response. HILDA offers a variety of variables measuring social support, including:  

1. Loneliness (Agreement with 'I often feel very lonely')
2. Friendless (Agreement with 'I don't have anyone to confide in')
3. Unsupported (Agreement with 'I have no one to lean on in times of trouble')
4. Helpless (Agreement with 'I often need help from people but can't get it')
5. Helped (Agremeent with 'When I need someone to help me I can usually find someone')  

We also examined _health_ (satisfaction with health) and _happiness_ (sum of SF-36 item 9).  

Other variables which may be included in the future include "feeling part of your local community" (`losatlc`), satisfaction with job security (`jbmssec`), and household net worth (`hwnwip`). See the [HILDA data dictionary](https://www.online.fbe.unimelb.edu.au/HILDAodd/srchVarnameUsingCategoriesCrossWave.aspx) for explanations of these variables.  

The time-varying dependencies among the social support variables over sixteen years were determined in a **Bayesian network**. Bayesian networks are a _graphical_ model which represent the joint probability distribution between the variables of interest. Graphical models represent variables of interest as nodes and the associations between them as arcs or edges. In a Bayesian network, the edges represent the conditional probability of a node given it's parent. A node may be conditionally dependent on a number of parent nodes, and also be associated with a number of child nodes. That is, a node may be a child of an upstream variable and a parent of a downstream variable. ***Dynamic* Bayesian networks** extend this fundamental idea to variables that are represented by nodes over time (i.e., a single variable measured at successive time points). When multiple variables are included, the result is a multivariate time-series. Within the multivariate time series, variables at the same time point cannot influence each other, or nodes at earlier time points; this guarantees the acyclicity of the graph (the Causal Markov Assumption). The unidirectional effect of time prevents feedback loops and is consistent with a causal interpretation of the resulting graph (Pearl, 2009).   

HILDA provides a multivariate time series for our variables of interest from 2001 to 2016. To learn the dynamic Bayesian network amoung our social support variables for each group (resilient, vulnerable and relieved), we used the developer version of the [EDISON](https://github.com/FrankD/EDISON) package. This version has updates to the matrix operations to significantly increase speed and improve performance.  

```
# Installing the developer version of EDISON:
devtools::install_github(repo='FrankD/EDISON', 
                         ref='MultipleTimeSeries', 
                         subdir='/Package/EDISON')
```

```{r import_subgroups, include=F}
# Import the different parametric response-groups identified in for-better-or-
# worse
#
# func to load the subgroups classified by the largest response component
find_subgroups <- function(parametric_fit) {
  
  coefs <- coef(parametric_fit)[[1]] # a 3D array 
  individualCoefs <- data.frame(
    "xwaveid" = rownames(coefs[, , 1]),
    "comp1" = unname(coefs[, 1, 2]),
    "comp2" = unname(coefs[, 1, 3]),
    "comp3" = unname(coefs[, 1, 4])
  )
  
  # Note: ranef() and fixef() can be used to extract the random and fixed effects
  # 
  # Find the maximum component and its sign
  individualCoefs$maxc <- max.col(abs(individualCoefs[, -1]))
  idx = individualCoefs$maxc+1
  individualCoefs$maxc <- individualCoefs$maxc*
    sign(as.numeric(individualCoefs[cbind(seq(length(idx)), idx)]))
  
  return(individualCoefs)
}

# event names
event_key <- c(
  `ledsc` = "widowed",
  `lefnw` = "bankruptcy",
  `lesep` = "divorce"
)

# import subgroups
subgroups <- map_dfr(
  .x = read_rds("data/parametric_fits.rds"), 
  .f = ~find_subgroups(.x), 
  .id = "eventcode"
) %>%
  mutate(eventcode = recode(eventcode, !!!event_key))

subgroups <- filter(subgroups, maxc %in% c(1, -1, 2)) %>%
  mutate(response = recode_factor(maxc, 
                                  `2` = "resilient",
                                  `-1` = "vulnerable", 
                                  `1` = "relieved", 
                                  .default = "undefined"))
```

```{r import_covariates, include=F}
# Import outcomes from for-better-or-worse
df <- read_rds("data/outcomes.rds") %>%
  rowwise() %>%
  mutate(year = which(letters == wave) + 2000) %>%
  ungroup()

# Import covariates from for-better-or-worse and bind with outcomes
covariates <- read_rds("data/covariates.rds") %>%
  select(-female) %>%
  bind_rows(df)
```

```{r create_inputs, eval=T}
# target variables
vars_of_interest <- c(
  lonely = "lssupvl", # Agreement with I often feel very lonely
  friend = "lssupac", # Agreement with I don't have anyone to confide in
  support = "lssuplt", # Agreement with I have no one to lean on in times of trouble
  helpless = "lssupnh", # Agremeent with I often need help from people but can't get it
  helped = "lssupsh", # Agreement with When I need someone to help me out I can usually find someone
  health = "losatyh", # Satisfaction with health
  happy = "gh9_sum"
  )

create_input <- function(.df, .event, .subgroup, .variables) {
  
  # filter the event of interest (e.g., "bankruptcy")
  df.filtered <- .df %>%
    filter(eventcode == .event) %>% 
    select(xwaveid, response) %>%
    left_join(filter(covariates, code %in% vars_of_interest)) %>%
    mutate(val = ifelse(val < 0, NA_real_, val)) %>%
    filter(response == .subgroup) %>%
    select(xwaveid, wave, code, val) %>%
    arrange(xwaveid, code, wave)
  
  # create a matrix of variables for EDISON
  matrixlist <- list()
  for (variable in .variables) {
    df.filtered %>%
      filter(code == variable) %>%
      spread(wave, val) %>%
      gather(wave, val, a:p) %>% 
      group_by(xwaveid) %>%
      fill(val, .direction = "downup") %>%
      mutate(sval = as.numeric(scale(val))) %>%
      ungroup() %>%
      mutate(sval = if_else(is.na(sval), 0, sval)) %>%
      dplyr::select(-val, -code) %>%
      spread(wave, sval) %>%
      column_to_rownames("xwaveid") %>% 
      as.matrix() -> mat
    matrixlist[[variable]] <- mat
  }
  
  input.array <- simplify2array(matrixlist)
  
  return(input.array)
}

estimate_model <- function(.input) {
  
  requireNamespace("EDISON")
  
  # Specify inference with no changepoints
  edison.options = EDISON::defaultOptions()
  edison.options$cp.fixed = TRUE
  
  start.time <- Sys.time()
  edison.result = EDISON::EDISON.run(.input, num.iter=50000, options=edison.options)
  end.time <- Sys.time()
  print(end.time - start.time)
  
  return(edison.result)
}

get_network <- function(.results, .variables, .threshold = 0.95) {
  adj.matrix <- EDISON::calculateEdgeProbabilities(.results)$probs.segs[[1]]
  colnames(adj.matrix) = rownames(adj.matrix) = names(.variables)
  network <- igraph::graph_from_adjacency_matrix(adj.matrix > .threshold, diag = F)
  return(network)
}
```

#### Bankruptcy

```{r bankruptcy}
thisevent = "bankruptcy"
```

```{r resilient_bankruptcy, include=F}
responsetype = "resilient"
resilient.input <- create_input(.df = subgroups,
                                .event = thisevent, 
                                .subgroup = responsetype, 
                                .variables = vars_of_interest)

# Calculate posterior probabilities of the edges in the network
resilient.results <- estimate_model(resilient.input)
```

```{r vulnerable_bankruptcy, include=F}
responsetype = "vulnerable"
vulnerable.input <- create_input(.df = subgroups,
                                 .event = thisevent, 
                                 .subgroup = responsetype, 
                                 .variables = vars_of_interest)

# Calculate posterior probabilities of the edges in the network
vulnerable.results <- estimate_model(vulnerable.input)
```

```{r relieved_bankruptcy, include=F}
responsetype = "relieved"
relieved.input <- create_input(.df = subgroups,
                               .event = thisevent, 
                               .subgroup = responsetype, 
                               .variables = vars_of_interest)

# Calculate posterior probabilities of the edges in the network
relieved.results <- estimate_model(relieved.input)
```

```{r save_bankruptcy}
write_rds(
  list(
    resilient.results = resilient.results, 
    vulnerable.results = vulnerable.results, 
    relieved.results = relieved.results),
  "results/bankruptcy.rds"
)
```

```{r plot_bankruptcy_networks, fig.width=9, fig.height=4, eval=T}
bankruptcy <- read_rds("results/bankruptcy.rds")

resilient.results  = bankruptcy[['resilient.results']]
resilient.network <- get_network(.results = resilient.results, 
                                 .variables = vars_of_interest)
resilient.degrees <- sort(igraph::degree(resilient.network), decreasing = T)

vulnerable.results = bankruptcy[['vulnerable.results']]
vulnerable.network <- get_network(.results = vulnerable.results, 
                                  .variables = vars_of_interest)
vulnerable.degrees <- sort(igraph::degree(vulnerable.network), decreasing = T)

relieved.results   = bankruptcy[['relieved.results']] 
relieved.network <- get_network(.results = relieved.results, 
                                .variables = vars_of_interest)
relieved.degrees <- sort(igraph::degree(relieved.network), decreasing = T)

par(mfrow=c(1,3))

node_colors = c(rep("lightsteelblue1", 5), "lightsteelblue3", "lightsteelblue3")

igraph::plot.igraph(resilient.network, 
     main = "Resilient network",
     vertex.size = 60, 
     edge.arrow.size=.5, 
     edge.arrow.mode=T,
     vertex.color=node_colors,
     vertex.frame.color="grey50",
     layout = igraph::layout_in_circle(resilient.network)
     )

igraph::plot.igraph(vulnerable.network, 
     main = paste("Vulnerable network"),
     vertex.size = 60, 
     edge.arrow.size=.5,
     edge.arrow.mode=T,
     vertex.color=node_colors,
     vertex.frame.color="grey50",
     layout = igraph::layout_in_circle(vulnerable.network)
     )

igraph::plot.igraph(relieved.network, 
     main = paste("Relieved network"),
     vertex.size = 60, 
     edge.arrow.size=.5, 
     edge.arrow.mode=T,
     vertex.color=node_colors,
     vertex.frame.color="grey50",
     layout = igraph::layout_in_circle(relieved.network)
     )
```

<br>

The maximum connectedness of the resilient graph was `r max(resilient.degrees)`, and the most interconnected variables (with the largest number of edges) were `r paste(names(resilient.degrees), resilient.degrees, sep = ":", collapse = ", ")`.  

The maximum connectedness of the vulnerable graph was `r max(vulnerable.degrees)`, and the most interconnected variables (with the largest number of edges) were `r paste(names(vulnerable.degrees), vulnerable.degrees, sep = ":", collapse = ", ")`.  

The maximum connectedness of the resilient graph was `r max(relieved.degrees)`, and the most interconnected variables (with the largest number of edges) were `r paste(names(relieved.degrees), relieved.degrees, sep = ":", collapse = ", ")`.  

<br>

#### Divorce

```{r divorce}
thisevent = "divorce"
```


```{r resilient_divorce, include=F}
responsetype = "resilient"
resilient.input <- create_input(.df = subgroups,
                                .event = thisevent, 
                                .subgroup = responsetype, 
                                .variables = vars_of_interest)

# Calculate posterior probabilities of the edges in the network
resilient.results <- estimate_model(resilient.input)
```

```{r vulnerable_divorce, include=F}
responsetype = "vulnerable"
vulnerable.input <- create_input(.df = subgroups,
                                 .event = thisevent, 
                                 .subgroup = responsetype, 
                                 .variables = vars_of_interest)

# Calculate posterior probabilities of the edges in the network
vulnerable.results <- estimate_model(vulnerable.input)
```

```{r relieved_divorce, include=F}
responsetype = "relieved"
relieved.input <- create_input(.df = subgroups,
                            .event = thisevent, 
                            .subgroup = responsetype, 
                            .variables = vars_of_interest)

# Calculate posterior probabilities of the edges in the network
relieved.results <- estimate_model(relieved.input)
```

```{r save_divorce}
write_rds(
  list(
    resilient.results = resilient.results, 
    vulnerable.results = vulnerable.results, 
    relieved.results = relieved.results),
  "results/divorce.rds"
)
```

```{r plot_divorce_networks, fig.width=9, fig.height=4, eval=T}
divorce <- read_rds("results/divorce.rds")

resilient.results  = divorce[['resilient.results']]
resilient.network <- get_network(.results = resilient.results, 
                                 .variables = vars_of_interest)
resilient.degrees <- sort(igraph::degree(resilient.network), decreasing = T)

vulnerable.results = divorce[['vulnerable.results']]
vulnerable.network <- get_network(.results = vulnerable.results, 
                                  .variables = vars_of_interest)
vulnerable.degrees <- sort(igraph::degree(vulnerable.network), decreasing = T)

relieved.results   = divorce[['relieved.results']] 
relieved.network <- get_network(.results = relieved.results, 
                                .variables = vars_of_interest)
relieved.degrees <- sort(igraph::degree(relieved.network), decreasing = T)

par(mfrow=c(1,3))

node_colors = c(rep("lightsteelblue1", 5), "lightsteelblue3", "lightsteelblue3")

igraph::plot.igraph(resilient.network, 
     main = "Resilient network",
     vertex.size = 60, 
     edge.arrow.size=.5, 
     edge.arrow.mode=T,
     vertex.color=node_colors,
     vertex.frame.color="grey50",
     layout = igraph::layout_in_circle(resilient.network)
     )

igraph::plot.igraph(vulnerable.network, 
     main = paste("Vulnerable network"),
     vertex.size = 60, 
     edge.arrow.size=.5,
     edge.arrow.mode=T,
     vertex.color=node_colors,
     vertex.frame.color="grey50",
     layout = igraph::layout_in_circle(vulnerable.network)
     )

igraph::plot.igraph(relieved.network, 
     main = paste("Relieved network"),
     vertex.size = 60, 
     edge.arrow.size=.5, 
     edge.arrow.mode=T,
     vertex.color=node_colors,
     vertex.frame.color="grey50",
     layout = igraph::layout_in_circle(relieved.network)
     )
```

<br>

The maximum connectedness of the resilient graph was `r max(resilient.degrees)`, and the most interconnected variables (with the largest number of edges) were `r paste(names(resilient.degrees), resilient.degrees, sep = ":", collapse = ", ")`.  

The maximum connectedness of the vulnerable graph was `r max(vulnerable.degrees)`, and the most interconnected variables (with the largest number of edges) were `r paste(names(vulnerable.degrees), vulnerable.degrees, sep = ":", collapse = ", ")`.  

The maximum connectedness of the resilient graph was `r max(relieved.degrees)`, and the most interconnected variables (with the largest number of edges) were `r paste(names(relieved.degrees), relieved.degrees, sep = ":", collapse = ", ")`.  

<br>

#### Death of partner

```{r widowed}
thisevent = "widowed"
```

```{r resilient_widowed, include=F}
responsetype = "resilient"
resilient.input <- create_input(.df = subgroups,
                                .event = thisevent, 
                                .subgroup = responsetype, 
                                .variables = vars_of_interest)

# Calculate posterior probabilities of the edges in the network
resilient.results <- estimate_model(resilient.input)
```

```{r vulnerable_widowed, include=F}
responsetype = "vulnerable"
vulnerable.input <- create_input(.df = subgroups,
                                 .event = thisevent, 
                                 .subgroup = responsetype, 
                                 .variables = vars_of_interest)

# Calculate posterior probabilities of the edges in the network
vulnerable.results <- estimate_model(vulnerable.input)
```

```{r relieved_widowed, include=F}
responsetype = "relieved"
relieved.input <- create_input(.df = subgroups,
                            .event = thisevent, 
                            .subgroup = responsetype, 
                            .variables = vars_of_interest)

# Calculate posterior probabilities of the edges in the network
relieved.results <- estimate_model(relieved.input)
```

```{r save_widowed}
write_rds(
  list(
    resilient.results = resilient.results, 
    vulnerable.results = vulnerable.results, 
    relieved.results = relieved.results),
  "results/widowed.rds"
)
```

```{r plot_widowed_networks, fig.width=9, fig.height=4, eval=T}
widowed <- read_rds("results/widowed.rds")

resilient.results  = widowed[['resilient.results']]
resilient.network <- get_network(.results = resilient.results, 
                                 .variables = vars_of_interest)
resilient.degrees <- sort(igraph::degree(resilient.network), decreasing = T)

vulnerable.results = widowed[['vulnerable.results']]
vulnerable.network <- get_network(.results = vulnerable.results, 
                                  .variables = vars_of_interest)
vulnerable.degrees <- sort(igraph::degree(vulnerable.network), decreasing = T)

relieved.results   = widowed[['relieved.results']] 
relieved.network <- get_network(.results = relieved.results, 
                                .variables = vars_of_interest)
relieved.degrees <- sort(igraph::degree(relieved.network), decreasing = T)


par(mfrow=c(1,3))

node_colors = c(rep("lightsteelblue1", 5), "lightsteelblue3", "lightsteelblue3")

igraph::plot.igraph(resilient.network, 
     main = "Resilient network",
     vertex.size = 60, 
     edge.arrow.size=.5, 
     edge.arrow.mode=T,
     vertex.color=node_colors,
     vertex.frame.color="grey50",
     layout = igraph::layout_in_circle(resilient.network)
     )

igraph::plot.igraph(vulnerable.network, 
     main = paste("Vulnerable network"),
     vertex.size = 60, 
     edge.arrow.size=.5,
     edge.arrow.mode=T,
     vertex.color=node_colors,
     vertex.frame.color="grey50",
     layout = igraph::layout_in_circle(vulnerable.network)
     )

igraph::plot.igraph(relieved.network, 
     main = paste("Relieved network"),
     vertex.size = 60, 
     edge.arrow.size=.5, 
     edge.arrow.mode=T,
     vertex.color=node_colors,
     vertex.frame.color="grey50",
     layout = igraph::layout_in_circle(relieved.network)
     )
```

<br>

The maximum connectedness of the resilient graph was `r max(resilient.degrees)`, and the most interconnected variables (with the largest number of edges) were `r paste(names(resilient.degrees), resilient.degrees, sep = ":", collapse = ", ")`.  

The maximum connectedness of the vulnerable graph was `r max(vulnerable.degrees)`, and the most interconnected variables (with the largest number of edges) were `r paste(names(vulnerable.degrees), vulnerable.degrees, sep = ":", collapse = ", ")`.  

The maximum connectedness of the resilient graph was `r max(relieved.degrees)`, and the most interconnected variables (with the largest number of edges) were `r paste(names(relieved.degrees), relieved.degrees, sep = ":", collapse = ", ")`.  

<br>